Flaring star-planet interactions in Kepler and TESS

### 2023-06-08 src/tex/ms.tex
[showyourwork] overleaf sync

\subsection{Flaring star-planet interactions}
\label{sec:intro:fspi}
The absolute power of flaring star-planet interactions is difficult to constrain. We do not know how energetic planet-induced flares can be. The estimates for dissipated energy~\citep{lanza2018closeby} are in the ballpark of the regime of superflares observed by Kepler and TESS. However, on the one hand, the planet might be triggering flares prematurely, which would be of lower energy~\citep{loyd2023flares}, and end up below our detection threshold. On the other hand, the energy need not be dissipated continuously, particularly in the stretch-and-break mechanism proposed by~\citep{lanza2012starplanet}, so an individual flare's energy might be higher. We also do not know whether flares triggered by interactions with a companion would have different properties than intrinsic flares. We will assume that, since they would be caused by magnetic reconnection triggered by the planet in the stellar corona, they should still have the same physics as intrinsic flares.

@@

\begin{figure*}[ht!]
\script{paper_illustrative_sketch_of_system_geometries.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/system_geometries.png}
        \caption{
         Simplified magnetic star-planet interaction geometries.\textbf{ Panel a: Fully aligned system along the line of sight. }Magnetic, rotational, and orbital axis are aligned. The interaction footpoint moves along the black line, and is visible on the red line, i.e. during $50\%$ of the planet's orbit. Flares induced along this line will result in excess flares when the planet is in front of the star. \textbf{Panel b: Orbital misalignment.} The rotational and magnetic axis are aligned, but the planet is highly misaligned. The footpoint of interaction moves between the two horizontal black lines, and is still visible $50\%$ of the time, resulting in modulated flaring with orbital phase. However, the footpoint passes different latitudes on the star, which correspond to different dipole field strengths, which may influence the power of interaction. \textbf{Panel c: Inclined fully aligned system.} The red interaction footpoint line is visible $100\%$ of the time. No modulation due to SPI can be observed, even if flares are triggered by the planet. The sketch only shows the interaction footpoint region on the Northern hemisphere of the star, but they may appear on both.
        }
        \label{fig:sketch}
    \end{centering}
\end{figure*}

@@

\subsubsection{HIP 67522}
\label{sec:results:individualstars:hip67522}
HIP 67522 has one of the strongest expected SPI signals in our sample, and shows the clearest sign of flaring SPI at $>2\sigma$ level, albeit with only 6 flares in the sample. These flares are distributed across two Sectors in TESS. In Sector 11, two flares occur at orbital phases $\sim 0.61$ and $\sim 0.81$. Two years later, in Sector 38, four flares occur, but this time, all of them take place at orbital phases $0.00-0.05$, that is, within 10 hours after mid-transit.

HIP 67522 is a young Sun, currently contracting onto the main sequence. It is a Sco-Cen member (10-20 Myr old), which was discovered to host a close-in Jupiter, HIP 67522 b, in 2020~\citep{rizzuto2020tess}. Curiously, the planet is in close spin-orbit commensurability -- $P_{\rm rot}/P_{\rm orb}\approx 1/5$. Therefore, we cannot completely rule out that the observed periodicity is in fact a rotational periodicity, and not related to the planet. However, if we calculate the $p$-value of the AD test using the rotational period instead of the orbital period, we find HIP 67522 consistent with uniform flaring in time within $1\sigma$, which is likely due to the not sufficiently low uncertainty in rotation period. So we cannot rule out rotational modulation yet, although the clustering close to transit is suggestive.% Fortunately, HIP 67522 will be observed again with TESS in April 2023, which may resolve the puzzle.

@@

Magnetic star-planet interaction can only take place if the planet spends at least a part of its orbit in the sub-Alfv\'enic regime~(see Section~\ref{sec:intro:mspi}).

Fig.~\ref{fig:as}

The average Alfv\'en surface size ($\overline{AS}_{R}$) was estimated using the power law  (Eq.~\ref{AS_estimation}) established by \citet{Chebly2023}:

@@

In \cite{reiners2022magnetism} it was reported that for Sun-like stars with a total magnetic field strength of up to 2000~G, the longitudinal magnetic field derived from Stokes V spectropolarimetric observations ($B_{\rm V}$) accounts for $\sim$\,$10$\% of the total magnetic field. \textbf{add ref.} showed that this trend also holds for M-dwarfs with magnetic fields up to 4000~G. Therefore, as a first assumption, we consider that the longitudinal magnetic field of our stars would reach 15\% of the total magnetic field (i.e. $B_{\rm V} = 0.15B$), given that most systems in our sample have B~$>$~2000 G.

@@

\textcolor{blue}{what we think about it (E and J contributions, comments from others  welcome)
TAP 26 and KOI-12 appear sub-Alfv\'enic in our estimate. So it seems unlikely that the absence of interaction is caused by the planets orbiting outside the Alfv\'en radius.} \textcolor{red}{JDAG: Thats a good point. However, I think being inside the AS is a required but not sufficient condition for the type of SPI that you are looking for. Because you are searching for flares, you need reconnection + particle acceleration (which are the precursors/drivers of the flare emission observed on different wavelengths). While the coronal material is expected to provide the particles (at least in closed-field Active Regions -- if the reconnection takes place in open-field line regions, the plasma density there is usually low which would also yield weak EM signatures as those are typically a function of $\rho$), one of the difficulties I see is that in the star-planet system, the reconnection of the magnetic field is more likely to occur in the planet end rather than on the stellar end (simply because the AR field strength is probably larger than the planetary field). As such, the field topology will be more 'resistant' closer to the star than to the planet, facilitating the re-organization in the latter case compared to the former. Planetary outflows falling back to the star could help to trigger some instabilities, but those would depend on whether or not the planet is losing any significant mass (which we probably don't know for these targets). There's of course more to it than just field strength or accretion, with things like the torsion and helicity of the field among others (which is why the likelihood of a solar AR to host a flare is not only determined by field strength but also by its topology/complexity), but all of that is too complicated or even impossible to estimate here. The bottom line is that there are multiple reasons that could explain why those two stars that lie deep in the sub-Alfv\'enic regime show no signs of flaring SPI which, without a better characterization of each individual system + some 3D modelling, are very hard to ascertain.}

@@

\subsection{Viewing geometry}
\label{sec:discussion:viewing}
Modulated visibility of the magnetic interaction footpoint is crucial for the detection of flaring star-planet interactions in our analysis. The viewing geometry of the star-planet system is therefore important to consider as an alternative to, or explanation for, the intermittency discussed above. Assuming for the sake of the argument that the sub-planetary point is close to the magnetic interaction footpoint in longitude, two factors will determine whether the footpoint's visibility will be modulated with the planetary orbit or not: the orbital inclination and the footpoint's latitude relative to it.

@@

Applying both the stretch-and-break and Alfv\'en wing mechanisms for magnetic interaction, we found a tentative evidence trend between the expected power of interaction, and the presence of excess flaring in phase with the orbital period $P_{\rm orb}$. In particular, we found that there may be two branches at high expected power -- one where the measured excess flaring is correlated with the expected power, and one where excess flaring is absent. This may be, among others, an effect of the extent of the Alfv\'en surface, the intermittency in the interaction, or of the viewing geometry of the systems.

>>>>

\subsection{Flaring star-planet interactions}
\label{sec:intro:fspi}

\begin{figure*}[ht!]
\script{paper_illustrative_sketch_of_system_geometries.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/system_geometries.png}
        \caption{
         Simplified magnetic star-planet interaction (SPI) geometries.\textbf{ Panel a: Fully aligned system along the line of sight (LOS). }Magnetic, rotational, and orbital axis are aligned. The interaction footpoint moves along the black line, and is visible to the observer on the red line, i.e. during $50\%$ of the planet's orbit. Flares induced along this line will result in excess flares when the planet is in front of the star. \textbf{Panel b: Orbital misalignment.} Same a panel a., but the planet's orbit is misaligned. The footpoint of interaction is again visible $50\%$ of the time, resulting in modulated flaring with orbital phase. However, it passes different latitudes on the star, which correspond to different dipole field strengths and coronal regions, which may affect the power of interaction. \textbf{Panel c: Inclined fully aligned system.} The red interaction footpoint line is visible $100\%$ of the time. No modulation due to magnetic SPI can be observed, even if flares are triggered by the planet. The sketches only show the interaction on one hemisphere of the star, but they may appear on both.
        }
        \label{fig:sketch}
    \end{centering}
\end{figure*}

The absolute power of flaring star-planet interactions is difficult to constrain. We do not know how energetic planet-induced flares can be. The estimates for dissipated energy~\citep{lanza2018closeby} are in the ballpark of the regime of superflares observed by Kepler and TESS. However, on the one hand, the planet might be triggering flares prematurely, which would be of lower energy~\citep{loyd2023flares}, and end up below our detection threshold. On the other hand, the energy need not be dissipated continuously, particularly in the stretch-and-break mechanism proposed by~\citep{lanza2012starplanet}, so an individual flare's energy might be higher. We also do not know whether flares triggered by interactions with a companion would have different properties than intrinsic flares. We will assume that, since they would be caused by magnetic reconnection triggered by the planet in the stellar corona, they should still have the same physics as intrinsic flares.

@@

@@

\subsubsection{HIP 67522}
\label{sec:results:individualstars:hip67522}
HIP 67522 has one of the strongest expected SPI signals in our sample, and shows the clearest sign of flaring SPI at $>2.5\sigma$ level, with 12 flares in the sample, distributed across three Sectors in TESS. In our original sample from July 2022, HIP 67522 was observed for two Sectors with a total of 6 flares, which yielded a $p$-value of $0.015$. In June 2023, observations from Sector 64 were released, which we used to follow up on this interesting target, increasing the number of flares by another 6, reducing the $p$-value further to $0.0057$.

HIP 67522 is a young Sun, currently contracting onto the main sequence. It is a Sco-Cen member (10-20 Myr old), which was discovered to host a close-in Jupiter, HIP 67522 b, in 2020~\citep{rizzuto2020tess}. Curiously, the planet is in close spin-orbit commensurability -- $P_{\rm rot}/P_{\rm orb}\approx 1/5$. Therefore, we cannot completely rule out that the observed periodicity is in fact a rotational periodicity, and not related to the planet. However, if we calculate the $p$-value of the AD test using the rotational period instead of the orbital period, we find HIP 67522 deviates from uniform flaring in time with $p=0.18$. So we cannot rule rotational modulation completely, since the rotational period is very uncertain relative to the total observing baseline, and the clustering close to transit is suggestive.% Fortunately, HIP 67522 will be observed again with TESS in April 2023, which may resolve the puzzle.

@@

Magnetic star-planet interaction can only take place if the planet spends at least a part of its orbit in the sub-Alfv\'enic regime~(see Section~\ref{sec:intro:mspi}). We estimated the average Alfv\'en surface size ($\overline{AS}_{R}$) using the power law established by \citet{Chebly2023}:

@@

In \cite{reiners2022magnetism} it was reported that for Sun-like stars with a total magnetic field strength of up to 2000~G, the longitudinal magnetic field derived from Stokes V spectropolarimetric observations ($B_{\rm V}$) accounts for $\sim$\,$10$\% of the total magnetic field. \cite{kochukhov2020hidden} showed that this trend also holds for M-dwarfs with magnetic fields up to 4000~G. In this case, however, the recovered longitudinal magnetic field accounts for about 20\% of the total magnetic field. Moreover, our star sample covers a wide range of magnetic fields, the weakest being 100~G and the strongest 3800~G. Therefore, as a first assumption, we consider that the longitudinal magnetic field of our stars would reach 15\% of the total magnetic field (i.e. $B_{\rm V} = 0.15B$), given that most systems in our sample have B~$>$~2000 G.

@@

We show the difference between star-planet distance and $\overline{AS}_{R}$ in Fig.~\ref{fig:as}. HIP 67522 b falls most deeply into the sub-Alfv\'enic regime, consistent with the strongest signal of magnetic SPI in our sample. TAP 26 and KOI-12 also appear sub-Alfv\'enic, despite absent signal of magnetic SPI. From this follows that the absence of interaction is not caused by the planets orbiting outside the Alfv\'en radius. location inside $\overline{AS}_{R}$ is a required but not sufficient condition for magnetic SPI, as we will see in the following Sections. %\textcolor{red}{JDAG: Thats a good point. However, I think being inside the AS is a required but not sufficient condition for the type of SPI that you are looking for. Because you are searching for flares, you need reconnection + particle acceleration (which are the precursors/drivers of the flare emission observed on different wavelengths). While the coronal material is expected to provide the particles (at least in closed-field Active Regions -- if the reconnection takes place in open-field line regions, the plasma density there is usually low which would also yield weak EM signatures as those are typically a function of $\rho$), one of the difficulties I see is that in the star-planet system, the reconnection of the magnetic field is more likely to occur in the planet end rather than on the stellar end (simply because the AR field strength is probably larger than the planetary field). As such, the field topology will be more 'resistant' closer to the star than to the planet, facilitating the re-organization in the latter case compared to the former. Planetary outflows falling back to the star could help to trigger some instabilities, but those would depend on whether the planet is losing any significant mass (which we probably don't know for these targets). There's of course more to it than just field strength or accretion, with things like the torsion and helicity of the field among others (which is why the likelihood of a solar AR to host a flare is not only determined by field strength but also by its topology/complexity), but all of that is too complicated or even impossible to estimate here. The bottom line is that there are multiple reasons that could explain why those two stars that lie deep in the sub-Alfv\'enic regime show no signs of flaring SPI which, without a better characterization of each individual system + some 3D modelling, are very hard to ascertain.}

@@

\subsection{Viewing geometry}
\label{sec:discussion:viewing}
Modulated visibility of the magnetic interaction footpoint is crucial for the detection of flaring star-planet interactions in our analysis~(Fig.~\ref{fig:sketch}). The viewing geometry of the star-planet system is therefore important to consider as an alternative to, or explanation for, the intermittency discussed above. Assuming for the sake of the argument that the sub-planetary point is close to the magnetic interaction footpoint in longitude, two factors will determine whether the footpoint's visibility will be modulated with the planetary orbit or not: the orbital inclination and the footpoint's latitude relative to it.

@@

Applying both the stretch-and-break and Alfv\'en wing mechanisms for magnetic interaction, we found a tentative trend between the expected power of interaction, and the presence of excess flaring in phase with the orbital period $P_{\rm orb}$. In particular, we found that there may be two branches at high expected power -- one where the measured excess flaring is correlated with the expected power, and one where excess flaring is absent. While we can exclude the extent of the Alfv\'en surface as a reason, intermittency in the interaction, and the viewing geometry of the systems are candidates for explaining the branching.
